---
title: "Project2"
format: html
editor: visual
author: Ralph Church
---

### Imports
```{r}
library(readr)
library(dplyr)
library(ggplot2)
```

### Loading the data

Fields are also converted to factors where necessary.

```{r}
bank_train <- read_delim(
  "data/train.csv",
  delim = ";",
  show_col_types = FALSE
)

bank_test <- read_delim(
  "data/test.csv",
  delim = ";",
  show_col_types = FALSE
)

bank <- bind_rows(bank_train, bank_test)

# Number of rows and columns
dim(bank)

#convert column variables to factors
bank <- bank %>%
  mutate(
    across(
      c(
        job,
        marital,
        education,
        default,
        housing,
        loan,
        contact,
        month,
        poutcome,
        y
      ),
      as.factor
    )
  )

# Variable names and types
glimpse(bank)

# First few observations
head(bank)

# Summary statistics
summary(bank)
```

### One-way contigency table

This shows how many customers subscribed versus did not subscribe:
```{r}
table(bank$y)
```

### Two-way contingency table

Marital status versus whether the customer subscribed:
```{r}
table(bank$marital, bank$y)
```

### Numerical Summaries

```{r}
bank %>%
  group_by(y) %>%
  summarize(
    mean_age = mean(age),
    median_age = median(age),
    sd_age = sd(age),
    mean_balance = mean(balance),
    median_balance = median(balance),
    sd_balance = sd(balance),
    mean_duration = mean(duration),
    .groups = "drop"
  )
```

### Plot 1: Bar Chart

This bar chart shows the number of customers who did and did not subscribe to the term deposit.

```{r}
ggplot(bank, aes(x = y)) +
  geom_bar() +
  labs(
    title = "Term Deposit Subscription Counts",
    x = "Subscribed to Term Deposit",
    y = "Number of Customers"
  )
```

Most customers did not subscribe to the term deposit. Only a small portion of customers chose to subscribe.

### Plot 2: Side-by-side bar chart

This bar chart compares customers with and without a housing loan based on whether they subscribed to the term deposit.

```{r}
ggplot(bank, aes(x = housing, fill = y)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Subscription Status by Housing Loan",
    x = "Has Housing Loan",
    y = "Number of Customers",
    fill = "Subscribed"
  )
```

Most customers did not subscribe, regardless of housing loan status. Customers without a housing loan appear slightly more likely to subscribe than those with a housing loan.

### Plot 3: Histogram

This histogram shows the distribution of customer ages. The bars are colored based on whether the customer subscribed to the term deposit.

```{r}
ggplot(bank, aes(x = age, fill = y)) +
  geom_histogram(binwidth = 5, color = "white") +
  labs(
    title = "Age Distribution by Subscription Status",
    x = "Age",
    y = "Number of Customers",
    fill = "Subscribed"
  )
```

Most customers are between 30 and 50 years old. Across all age groups, more customers did not subscribe than those who did.

### Plot 4: Scatter plot

This scatter plot shows the relationship between a customer's age and account balance. The points are colored by whether the customer subscribed to the term deposit.

```{r}
ggplot(bank, aes(x = age, y = balance, color = y)) +
  geom_point(alpha = 0.4) +
  labs(
    title = "Age vs. Account Balance by Subscription Status",
    x = "Age",
    y = "Account Balance",
    color = "Subscribed"
  )
```

There is no strong relationship between age and account balance. Most customers have relatively low balances, regardless of whether they subscribed, although a few customers have much higher balances.

### Plot 5: Box Plot

This box plot compares the distribution of account balances for customers who did and did not subscribe to a term deposit. It helps identify differences in the median, spread, and potential outliers between the two groups.

```{r}
ggplot(bank, aes(x = y, y = balance, fill = y)) +
  geom_boxplot() +
  labs(
    title = "Account Balance by Subscription Status",
    x = "Subscribed to Term Deposit",
    y = "Account Balance",
    fill = "Subscribed"
  )
```

The box plot shows that customers who subscribed ("yes") generally have a slightly higher account balance than those who did not subscribe ("no"). Both groups contain many outliers, indicating that a small number of customers have unusually large account balances.

### Plot 6 not covered in class: Heatmap

This heatmap shows the number of customers in each job category who did and did not subscribe to the term deposit. Darker colors represent higher customer counts.

```{r}
heatmap_data <- bank %>%
  count(job, y)

ggplot(heatmap_data, aes(x = y, y = job, fill = n)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 3) +
  labs(
    title = "Heatmap of Job Type and Term Deposit Subscription",
    x = "Subscribed to Term Deposit",
    y = "Job Type",
    fill = "Customer Count"
  )
```

Most job groups have many more customers who did not subscribe than those who did. Blue-collar, management, and technician jobs have the largest number of customers overall.
